---
title: "ggplot2 Histogram Logo Fail"
author: "Matt Kmiecik"
date: "Last updated: `r format(Sys.Date(), format = '%d %B %Y')`"
---

```{r setup, include=FALSE}
#pixWidth <- 1200
#pixHeight <- .8 * pixWidth
#set_dpi <- 72
#inWidth <- pixWidth/set_dpi
#inHeight <- pixHeight/set_dpi
knitr::opts_chunk$set(fig.align = 'center', 
                      #fig.width = inWidth, 
                      #fig.height = inHeight, 
                      comment = '>')
```

When I first started creating this website (in WordPress), I thought it would be cool have have my own logo. Why not, right? I figured I'd go super nerdy and have a cool histogram plot with a density curve through it. I had just finished reading Hadley Wickham's ggplot2 [book](https://github.com/hadley/ggplot2-book) and I was excited to put some of those newly learned skills to use. 

First off, I'm not a design guy (if you can't tell already), but I had a vision of this plot and I just went with it. After making the basic design, I couldn't figure out how to get spaces between the bars in a ggplot2 histogram. After searching everywhere, the best advice I got was, ["Histograms don't have spaces between them, so why would you want them?"](https://groups.google.com/forum/#!topic/ggplot2/yTEm-OfsX18)

Then it dawned on me that I could plot two histograms, and then set their position to dodge and make one of the histograms invisible. So first, let's plot two histograms:

```{r two-histograms, warning=FALSE}
rm(list = ls())   # Clear workspace
library(ggplot2)  # Load ggplot2

set.seed(14)      # Set seed for reproducible example
samplesA <- 400   # 400 samples in the distribution

# Create dataframe for plotting
logoData <- data.frame(dist = rep(c('A', 'B'), each = samplesA), 
                       x = rnorm(2*samplesA))
# Logo plot
ggplot(logoData, aes(x, fill = dist, colour = dist)) +
  geom_histogram(aes(y = ..density..), 
                 position = 'dodge', 
                 bins = 13) +
  geom_density(alpha = 1/5) +
  scale_fill_manual(values = c('#1191e0', 'lightslategrey')) +
  scale_color_manual(values = c('#222222', '#222222')) +
  xlim(-4 , 4) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        legend.position = 'none',
        panel.background = element_rect(fill = '#eaeaea'))
```

Now, let's make one of the histograms invisible by setting its fill and color to NA, giving the illusion that there is only one histogram plotted:

```{r one-histogram, warning=FALSE}
ggplot(logoData, aes(x, fill = dist, colour = dist)) +
  geom_histogram(aes(y = ..density..), 
                 position = 'dodge', 
                 bins = 13) +
  geom_density(alpha = 1/5) +
  scale_fill_manual(values = c('#1191e0', NA)) +  # Setting fill to NA
  scale_color_manual(values = c('#222222', NA)) + # Setting color to NA
  xlim(-4 , 4) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        legend.position = 'none',
        panel.background = element_rect(fill = '#eaeaea'))
```

Now we are left with a single histogram density plot with spaces between the bars. However, when I tried to make this my website logo, I couldn't figure out a way to get a high resolution image at 100 x 100 pixels. Instead, my brother made me a sweet bar graph with hockey sticks through it in Photoshop! Either way, through this journey I was able to come up with a solution to having spaces between histogram bars in ggplot2. Hopefully this is useful to someone!

